---
title: "Midterm"
author: "Jacob Thielemier"
date: "2025-03-16"
output: pdf_document
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = NA,
  results = "markup"
)
```

## Question 1

-   The plot below shows pretty consistent behavior between different time points. Most of the subject follow a similar rise and then fall pattern.

-   Most subjects peak at a time period of 25. Several subjects follow each other by rising and falling at the same time periods. But a few outliers exists that show sudden deviations near 10 and 30.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)

# Read the EEG data
eeg_data <- read_csv("EEG.csv")

# Rename the first column to 'Time' (as it represents the time points)
colnames(eeg_data)[1] <- "Time"

# Reshape the data: each remaining column is a subject.
eeg_long <- eeg_data %>%
  pivot_longer(
    cols = -Time,         # All columns except Time are subject columns
    names_to = "Subject", # Create a new column named "Subject"
    values_to = "EEG_Value"  # Values will be stored in "EEG_Value"
  ) %>%
  mutate(Time = as.numeric(Time))  # Ensure Time is numeric

# Plot EEG time series for each subject
ggplot(eeg_long, aes(x = Time, y = EEG_Value, color = Subject)) +
  geom_line() +
  labs(
    title = "EEG Time Series by Subject",
    x = "Time Point",
    y = "EEG Value",
    color = "Subject"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend if too cluttered

```

## Question 2

-   The heatmap below shows that there is clustering among the subjects.

    -   One cluster in the top-left where the subjects are relatively genetically similar. The lighter yellow shades mean smaller distances.

    -   Another cluster in the bottom-right is darker red. Meaning those are more similar withing their own group, but also more distant from the other cluster.

-   The overall distance between the two cluster is far that suggests there is a high genetic dissimilarity between the subjects in the two clusters.

```{r}
library(reshape2)
library(pheatmap)

snp_data <- read_csv("SNP.csv")

snp_matrix <- snp_data %>%
  select(-1) %>%        # remove the first column (Subject ID)
  as.matrix()

# Compute Euclidean distance matrix
dist_matrix <- dist(snp_matrix, method = "euclidean")

# Convert distance object to a matrix
dist_matrix_mat <- as.matrix(dist_matrix)

# Plot the heatmap
pheatmap(
  dist_matrix_mat,
  main = "Genetic Similarity (Euclidean Distance) Heatmap",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean"
)
```

## Question 3

```{r}
library(FactoMineR)
library(knitr)

# Prepare the EEG data:
# Assume 'eeg_data' is in wide format with the first column as 'Time'
# and the remaining columns as subjects.
# Convert to a matrix and transpose so that rows represent subjects.
eeg_matrix <- as.matrix(eeg_data[,-1])
eeg_subjects <- t(eeg_matrix)

# Prepare the SNP data:
# Assume 'snp_matrix' is already a matrix with subjects as rows and SNPs as columns.
snp_subjects <- snp_matrix

# Compute the RV coefficient using FactoMineR's coeffRV function.
rv_result <- coeffRV(eeg_subjects, snp_subjects)

# Manually create a table with the desired values.
rv_table <- data.frame(
  Metric = c("RV", "RV Std", "Mean", "Variance", "Skewness", "p-value"),
  Value = c(
    round(rv_result$rv, 3),
    round(rv_result$rvstd, 3),
    round(rv_result$mean, 3),
    round(rv_result$variance, 5),
    round(rv_result$skewness, 3),
    format(rv_result$p.value, scientific = TRUE, digits = 3)
  ),
  stringsAsFactors = FALSE
)

# Print the table neatly using kable
kable(rv_table, 
      caption = "Summary of RV Coefficient Results from FactoMineR",
      align = "c")

```

## Question 4

```{r}
# Calculate the mean EEG value for each subject (ignoring the Time column)
eeg_means <- colMeans(eeg_data[,-1])

# Plot a histogram of the EEG means
hist(eeg_means, 
     breaks = 20,
     main = "Histogram of Subject EEG Means", 
     xlab = "Mean EEG Value", 
     col = "lightblue", 
     border = "white")
```

## Question 5

```{r}
library(VCA)
library(knitr)

# Assuming your long-format EEG data is stored in 'eeg_long'
long_eeg_data <- as.data.frame(eeg_long)

# Fit the one-way random effects model using anovaVCA
model_vca_full <- anovaVCA(EEG_Value ~ Subject, long_eeg_data)

# Print the model summary (for reference)
print(model_vca_full)

```

## Question 6

-   Based on the Scree plot below we see the elbow takes shape at the second PCA. We should only use two components.

```{r}
# Assuming the SNP data is loaded into 'snp_data'
# Remove the first column (subject identifier) and convert the rest to a matrix
snp_matrix <- as.matrix(snp_data[,-1])

# Perform PCA on the SNP matrix with scaling
pca_result <- prcomp(snp_matrix, scale. = TRUE)

# Extract the proportion of variance explained by each principal component
pve <- summary(pca_result)$importance[2, ]

# Produce the screeplot
plot(pve, type = "b",
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Screeplot of SNP Data")
```

## Question 7

-   There is strong evidence (p < 0.0001) that the EEG means are associated with PC1. However, PC2 is not significantly associated with EEG means (p = 0.729). This tell me that at least one genetic principal component (PC1) has a clear relationship with the EEG phenotype.

-   The adjusted R-squared is 0.396, meaning that 39.6% of the total variance in EEG means is explained by these genetic principal components.

```{r}
# Calculate the EEG means for each subject
eeg_means <- colMeans(eeg_data[,-1])

# Prepare the data frame with EEG means and selected PCs
pcs_selected <- pca_result$x[, 1:2]  # first two PCs selected based on screeplot
regression_df <- data.frame(
  EEG_Mean = eeg_means,
  PC1 = pcs_selected[, 1],
  PC2 = pcs_selected[, 2]
)

# Fit linear regression model
lm_fit <- lm(EEG_Mean ~ PC1 + PC2, data = regression_df)

# Extract important regression results
regression_summary <- summary(lm_fit)

# Create a clean data frame to display coefficients and p-values
regression_table <- data.frame(
  Predictor = rownames(regression_summary$coefficients),
  Estimate = round(regression_summary$coefficients[, 1], 4),
  `Std. Error` = round(regression_summary$coefficients[, 2], 4),
  `t-value` = round(regression_summary$coefficients[, 3], 3),
  `p-value` = round(regression_summary$coefficients[, 4], 4),
  check.names = FALSE
)

# Display the table using kable
library(knitr)
kable(regression_table, 
      caption = "Linear Regression of EEG Means on Genetic Principal Components")
```

## Question 8







